---
title: "lab6"
author: "Abeoseh Flemister"
date: "2026-04-09"
output: html_document
---

Read in data

```{r}
temp <- tempfile()
download.file("http://afodor.github.io/classes/stats2015/longitdunalRNASeqData.zip", temp)

counts <- read.table(unz(temp, "nc101_scaff_dataCounts.txt"),  sep = "\t", header = TRUE, row.names = 1)

annotations <- read.table(unz(temp, "nc101_Annotations.txt"),  sep = "\t", header = TRUE, row.names = 1)

unlink(temp)

print(head(counts))

print(head(annotations))
```

Remove rare genes and normalize

```{r cars}
counts <- counts[ apply( counts, 1, median)> 5,]

CountsNorm <- counts
for ( i in 1:ncol(counts))
{
	colSum = sum(counts[,i])
	CountsNorm[,i] = CountsNorm[,i]/colSum
}

```

(The first 3 columns are “day 2”, the next 3 columns are “week 12” and the last 5 are “week 18” (even though they say w20)). 

(A) For each row in the spreadsheet, perform a one-way ANOVA with categories “day 2”, “week 12” and “week 18”. Plot out the histogram of all p-values. How many genes are significant at a BH FDR-corrected 0.05 threshold.

```{r}

categories <- c(rep("D2",3), rep("W12",3), rep("W20",5))
categories <- factor(categories)
pvals <- vector(length = nrow(CountsNorm) )


for (i in 1:nrow(CountsNorm)){
  
  linear_reg <- lm( as.numeric(CountsNorm[i,]) ~ categories )
  
  pvals[i] <- anova(linear_reg)[["Pr(>F)"]][1]
  
  }


hist(pvals)

sig = p.adjust(pvals, method = "BH")
print(paste(sum(sig < 0.05),"p-vlaues are significant after benjamin hochberg correction"))

most_sig1 = row.names(CountsNorm)[ match( min(sig), sig ) ]
```

```{r}
row.names(CountsNorm)[ match( min(sig), sig ) ]
```

(A) Next make an ANOVA as a linear regression as a function of time (so 2 days, 86 days and 128 days).  Plot out the histogram of all p-values.  How many genes are significant at a BH FDR-corrected 0.05 threshold.

```{r}

days <- c(rep(2,3), rep(86,3), rep(128,5))
pvals <- vector(length = nrow(CountsNorm) )


for (i in 1:nrow(CountsNorm)){
  
  linear_reg <- lm( as.numeric(CountsNorm[i,]) ~ days )
  
  pvals[i] <- anova(linear_reg)[["Pr(>F)"]][1]
  
  }


hist(pvals)

sig = p.adjust(pvals, method = "BH")
print(paste(sum(sig < 0.05),"p-vlaues are significant after benjamin hochberg correction"))

most_sig2 = row.names(CountsNorm)[ match( min(sig), sig ) ]
```

(C) Finally, for each row in the spreadsheet perform an ANVOA comparing the three-parameter model from (A) and the two parameter model from (B).  Plot out the histogram of all p-values.  For how many genes is there a significant difference between these two models at a BH FDR-corrected threshold.

```{r}
x <- c(1,1,4,4,5,7,8,10)
y = c("a", "a", "b", )
x[which(x == 4)]

```

```{r}

## make the reduced model the linear model.
categories <- c(rep("D2",3), rep("W12",3), rep("W20",5))
categories <- factor(categories)

days <- c(rep(2,3), rep(86,3), rep(128,5))


pvals <- vector(length = nrow(CountsNorm) )


for (i in 1:nrow(CountsNorm)){
  
  
  FullModel <- lm( as.numeric(CountsNorm[i,]) ~ categories )
  FullResiduals <- sum( residuals(FullModel) ^ 2 )
  
  ReducedModel <- lm( as.numeric(CountsNorm[i,]) ~ days )
  ReducedResiduals <- sum( residuals(ReducedModel) ^ 2 )
  
  ## df  = # data points - # params 
  ## df in reduced = 11 - 2 = 9
  ## df in full = 11 - 3 = 8
  
  F_value <- ( (ReducedResiduals - FullResiduals) / (9 - 8) ) / (FullResiduals - 8)
  
  pvals[i] <- pf(F_value, (9-8), 8, lower.tail = F)
  }


hist(pvals)

sig = sum((p.adjust(pvals, method = "BH")) < 0.05)
print(paste(sig,"p-vlaues are significant after benjamin hochberg correction"))
```

```{r}

length(CountsNorm[i,])

linear_reg <- lm( as.numeric(CountsNorm[i,]) ~ categories, x = TRUE)

linear_reg$x
```
