---
title: "lab4"
author: "Abeoseh Flemister"
date: "2026-02-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 

```{r cars}
annotation <- read.table("./longitdunalRNASeqData/nc101_Annotations.txt", sep = "\t", header = TRUE, row.names = 1)
datacounts <- read.table("./longitdunalRNASeqData/nc101_scaff_dataCounts.txt", sep = "\t", header = TRUE, row.names = 1)
```

## (2) On a log10-log10 scale, show a plot of the counts for the two samples “D2_01” and “D2_02”.

Qualitatively, do the biological replicates appear to have similar patterns of gene expression?

**Qualitatively the biological replicates have similar patterns of gene expression since we can draw a roughly straight regression line.**

```{r pressure, echo=FALSE}

plot(log10(datacounts$D2_01), log10(datacounts$D2_02))
```

## (3) On a log10-log10 scale, plot the variance of all genes (across all genes) vs. the mean (across all genes) with a red line on your graph representing the identity line. Does the mean equal the variance for these samples?

(The commands apply(myT, 1, mean) and apply(myT, 1, var) will be of some use).

**The mean is not exactly equal to the variance which we can see through the deviations from the regression line.**

```{r}
## If there are all 0s in a row, log10 will be undefined 
datacounts_filtered = datacounts[apply(datacounts, 1, sum) != 0,]
means = log10(apply(datacounts_filtered, 1, mean))
vars = log10(apply(datacounts_filtered, 1, var))

plot(means, vars)
abline(summary(lm(vars ~ means))$coefficients[1,1], summary(lm(vars ~ means))$coefficients[2,1], col = "red")

```

## (4) Consider the first gene in the spreadsheet (e.g. NC101_00003). Make a two by two contingency table:

|                             | Sequences in D2_01 | Sequences in D2_02 |
|-----------------------------|--------------------|--------------------|
| Assigned to NC101_00003     |                    |                    |
| Not assigned to NC101_00003 |                    |                    |

```{r}
library(dplyr)

is_NC101_00003 = datacounts %>% select(D2_01, D2_02) %>% filter(row.names(datacounts) == "NC101_00003")

not_NC101_00003 = datacounts %>% select(D2_01, D2_02) %>% filter(row.names(datacounts) != "NC101_00003") %>% apply(2, sum)

fisher.test(rbind(is_NC101_00003, not_NC101_00003))
```

## (5) Now generate a p-value for all the genes in the spreadsheet from the Fisher test.  Plot out those p-values in a histogram.

-   Are they uniformly distributed? **The p-values are not normally distributed**
-   Would you expect them to be? **If I were to accept the null hypothesis I would expect the p-values to be uniformally distributed**
-   Are the p-values more significant, less significant or what we would expect under a uniform distribution? **Since the p-values are not uniformally distributed they are different than what would expect if we were to accept the null hypothesis. Specifically, the low p-values are from cases where the assumptions of the hypergeometric distribution are violated and the high p-values are the from the low abundance genes.**

```{r}
genes = row.names(datacounts)
unfilteredpvals = vector(length = length(genes))
i = 1

for( gene in genes){
  
  is_gene = datacounts %>% select(D2_01, D2_02) %>% filter(row.names(datacounts) == gene)
  
  not_gene = datacounts %>% select(D2_01, D2_02) %>% filter(row.names(datacounts) != gene) %>% apply(2, sum)
  
  unfilteredpvals[i] = fisher.test(rbind(is_gene, not_gene))$p.value
  
  i = i + 1
}
hist(unfilteredpvals, plot = TRUE, breaks = 10)
```

-   How does the p-value distribution change if you remove low abundance genes (with for example myT \<- myT[ (myT\$D2_01 + myT\$D2_02 \> 50),]
    -   Are they uniformly distributed? **The p-values are still not normally distributed**
    -   Would you expect them to be? **If I were to accept the null hypothesis I would expect the p-values to be uniformally distributed**
    -   Are the p-values more significant, less significant or what we would expect under a uniform distribution? **Since the p-values are still not uniformally distributed they are different than what would expect if we were to accept the null hypothesis. Specifically, the low p-values are from cases where the assumptions of hypergeometric distribution the are violated.**

```{r}
datacounts_filtered = datacounts[ (datacounts$D2_01 + datacounts$D2_02 > 50),] 
genes = row.names(datacounts_filtered)
filteredpvals = vector(length = length(genes))
i = 1

for( gene in genes){
  
  is_gene = datacounts_filtered %>% select(D2_01, D2_02) %>% filter(row.names(datacounts_filtered) == gene)
  
  not_gene = datacounts_filtered %>% select(D2_01, D2_02) %>% filter(row.names(datacounts_filtered) != gene) %>% apply(2, sum)
  
  filteredpvals[i] = fisher.test(rbind(is_gene, not_gene))$p.value
  
  i = i + 1
}


hist(filteredpvals, plot = TRUE, breaks = 10)
```

## (6) Add 1 to every value in the table ( with something like myT = myT + 1 ).

-   This is called adding a pseudo-count.

-   Now consider the first gene (NC101_00003 ) again.

-   From the first experiment, calculate expected frequency = p = (\# Assigned to NC101_00003 in D2_01)/total \# of sequences in D2_01)

```{r}
pseudocount_datacounts <- datacounts %>% select(D2_01, D2_02) + 1

is_NC101_00003 = pseudocount_datacounts %>% select(D2_01) %>% filter(row.names(datacounts) == "NC101_00003") %>% sum()

total_seqs_d1 = pseudocount_datacounts %>% select(D2_01) %>% sum()

expected_frequency = pseudocount_datacounts %>% select(D2_01) %>% filter(row.names(datacounts) == "NC101_00003") / total_seqs_d1

total_seqs_d2 = pseudocount_datacounts %>% select(D2_02) %>% sum()

poisson.test(is_NC101_00003, total_seqs_d2, expected_frequency$D2_01)


```

## (7) Repeat the calculation in (6) for every gene in the spreadsheet.  Graph these p-values against the p-values produced in (5) on a log10-log10 plot.  How well do they agree?

```{r}

pseudocount_datacounts <- datacounts %>% select(D2_01, D2_02) + 1

datacounts_filtered = datacounts[ (datacounts$D2_01 + datacounts$D2_02 > 50),] 
genes = row.names(pseudocount_datacounts)
poissonpvals = vector(length = length(genes))
i = 1

for (gene in genes){

  is_gene = pseudocount_datacounts %>% select(D2_01) %>% filter(row.names(datacounts) == gene) %>% sum()
  
  total_seqs_d1 = pseudocount_datacounts %>% select(D2_01) %>% sum()
  
  expected_frequency = pseudocount_datacounts %>% select(D2_01) %>% filter(row.names(datacounts) == gene) / total_seqs_d1
  
  total_seqs_d2 = pseudocount_datacounts %>% select(D2_02) %>% sum()
  
  poissonpvals[i] = poisson.test(is_gene, total_seqs_d2, expected_frequency$D2_01)$p.value
  
  i = i + 1

}

hist(poissonpvals, plot = TRUE, breaks = 10)
```

```{r}
plot(log10(poissonpvals), log10(unfilteredpvals))
```

**There is very little agreement between the two tests. However, the assumptions of both tests were violated so the p-values cannot be trusted. The poisson assumes the mean equals the variance which is not true in our dataset.**
