---
title: "Lab 1"
output: html_document
date: "2026-01-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Use the mean and sd functions to calculate the mean and standard deviation of the following numbers { 2.3, 4.3 ,1.2 , 3.4, 8.3, 12.2 }

```{r set1}
set1 <- c(2.3,  4.3 ,1.2 , 3.4, 8.3, 12.2)

print(paste("mean:", mean(set1), "std:", sd(set1)))
```

## Question 2

Use mean and sd in R to calculate the average and sd of { 4,5,6 …. 678,679,680 } (the set of every positive integer between 4 and 680)

```{r set2}
set2 <- 4:680
print(paste("mean:", mean(set2), "std:", sd(set2)))

```

## Question 3

Suppose you are given GC content for sequences from a human genome with values for many sequences: `gc <- runif(1000, min = 0.2, max = 0.8)`

1.  Plot a histogram of gc.
2.  Compute the mean and standard deviation.
3.  Does this look like a reasonable sequencing dataset? Why or why not?

```{r set3}
set.seed(0)
gc <- runif(1000, min = 0.2, max = 0.8) 
hist(gc)
print(paste("mean", mean(gc), "sd", sd(gc)))
```

This is not a reasonable sequencing dataset. For a single species (human), I would expect them to have relatively similar gc content. I would not expect a wide variety of gc contents across individuals. Instead, I would expect a normal distribution where most people have a similar gc content.
